Predicting Success of Motivational Interviews with Recurrent Neural Network and Probabilistic Model for Patient-Provider Communication Sequences

### 2017-08-30 paper-amia-summit-2018/amia_summit_2018.tex
update data collection, evaluation metrics and conclusion
update methods

Our proposed prediction method consists of two steps. In the first step, we model successful and unsuccessful patient-provider interactions using first- and second-order Markov Chain (MC) and Hidden Markov Model (HMM), which are popular probabilistic models for discrete observation sequences with finite vocabulary. Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence modeling \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk}.

@@

If $p(S\rightarrow CML+) > \delta $, where $\delta$ is an empirically determined threshold, the interview is predicted to result in positive change talk or commitment language. Tuning the proposed method on the training data involves experimentally determining the value of $\delta$, which results in the highest prediction F-measure with lower bias. Figure~\ref{fig:delta} illustrates the performance of Markov Chain-based prediction method by varying the value of $\delta$. As can be seen from Figure~\ref{fig:delta}, F-measure varies insignificantly, when $\delta $ is between -5 to -1, due to possible bias. However, F-score sharply decreases after $\delta = -0.9$. Thus, -0.9 is the optimal value of $\delta $, where F-measure is the highest with lower bias.

@@

We exclude the structure parameters M and N to designate HMM using a compact notation: $\lambda = (A, B, \pi)$. The key difference between HMM and MC is that HMM requires specifying the number of hidden states as a model parameter and then the model deduces a sequence of hidden states that best explains the observations along with state transition probabilities and distributions of observation symbols for each state. The Baum-Welch algorithm was used to estimate the parameters of HMMs for successful and unsuccessful interviews using the corresponding training set, while the Viterbi algorithm was used to determine the most likely sequence of hidden states for a given sequence of observations. After assignment of hidden states, the log-likelihood of successful outcome can be estimated using Eq.~\ref{eq:class}. As in the case of MC, we also experimented with second-order HMM.

@@

\section*{Results}
Experimental evaluation of the proposed method is conducted on both natural sequences and alternating sequences.

@@

\subsection*{\textit{Impact of hidden states}}
While MC is a non-parametric method, HMM has an important parameter, the number of hidden states. A natural question is how does this parameter affect the performance of HMM-based prediction  method. Figure~\ref{fig:hidden-states} provides an answer to this question by illustrating the performance of the proposed method using HMMs of different order in terms of F-measure on different types of sequences. As follows from this Figure, in general, the number of states does not significantly impact the classification performance. All four models show their best performance when the number of states is between 4 and 8.

Tables~\ref{tab:emission_matrix_s} and \ref{tab:emission_matrix_u} show patient-provider communication (PPC) codes with the highest probability in emission probability distributions associated with different states of HMMs trained on successful and unsuccessful interviews.\\

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.70\textwidth]{figures/hidden-states.png}
    \caption{\textbf{Performance of the proposed method using HMMs of different order by varying the number of hidden states}.}
    \label{fig:hidden-states}
\end{figure}

\begin{table}[!htb]
\centering
\caption{PPC codes with the highest probability associated with different states of an HMM trained on successful interviews.}
\label{tab:emission_matrix_s}
  \begin{tabular}{|l|l|}
  \hline
   \textbf{State} & \textbf{Most likely PPC codes} \\\hline 
1 & SS, EA, R-CT+, OQ-ECT+, SUM-S, AF, R-CML+ \\\hline           
2 & LUP+, EA, OQ-ECT+, OQ-ECML+  \\\hline    
  \end{tabular}
\end{table}

\begin{table}[!htb]
\centering
\caption{PPC codes with the highest probability associated with different states of an HMM trained on unsuccessful interviews.}
\label{tab:emission_matrix_u}
  \begin{tabular}{|l|l|}
  \hline
   \textbf{State} & \textbf{Most likely PPC codes} \\\hline     
1 & OQ-TBN, EA, R-CT+, AR, SUM-S, RBA-S, G-INFO+, R-CML+, OQ-ECT- \\\hline
2 & LUP+, OQ-ECT+, G-INFO+, CQ-EF, CHT-, HUP-O \\\hline
  \end{tabular}
\end{table}

>>>>

Our proposed baseline prediction method consists of two steps. In the first step, we model successful and unsuccessful patient-provider interactions using first- and second-order Markov Chain (MC) and Hidden Markov Model (HMM), which are popular probabilistic models for discrete observation sequences with finite vocabulary. In the second step, we classified each test sequence based on the maximum likelihood of generating that sequence from each model. Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence modeling \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk}. The latest advances in deep learning technologis shows that deep learning, in particular RNNs, provide better results than conventional machine learning methods for the task of sequence classification. In our experiment, we also employed two especial variant of RNN models: long short-term memory (LSTM) and gated recurrent unit (GRU).

@@

If $p(S\rightarrow CML+) > 0 $, the interview is predicted to result in positive change talk or commitment language.

@@

We exclude the structure parameters M and N to designate HMM using a compact notation: $\lambda = (A, B, \pi)$. The key difference between HMM and MC is that HMM requires specifying the number of hidden states as a model parameter and then the model deduces a sequence of hidden states that best explains the observations along with state transition probabilities and distributions of observation symbols for each state. The Baum-Welch algorithm was used to estimate the parameters of HMMs for successful and unsuccessful interviews using the corresponding training set, while the Viterbi algorithm was used to determine the most likely sequence of hidden states for a given sequence of observations. After assignment of hidden states, the log-likelihood of successful outcome can be estimated using Eq.~\ref{eq:class}.

@@

\section*{Results}
Experimental evaluation of the proposed method is conducted on both under- and over-sampled sequences.

@@

%\subsection*{\textit{Impact of hidden states}}
%While MC is a non-parametric method, HMM has an important parameter, the number of hidden states. A natural question is how does this parameter affect the performance of HMM-based prediction  method. Figure~\ref{fig:hidden-states} provides an answer to this question by illustrating the performance of the proposed method using HMMs of different order in terms of F-measure on different types of sequences. As follows from this Figure, in general, the number of states does not significantly impact the classification performance. All four models show their best performance when the number of states is between 4 and 8. 
%
%Tables~\ref{tab:emission_matrix_s} and \ref{tab:emission_matrix_u} show patient-provider communication (PPC) codes with the highest probability in emission probability distributions associated with different states of HMMs trained on successful and unsuccessful interviews.\\
%
%\begin{figure}[!htb]
%    \centering
%    \includegraphics[width=0.70\textwidth]{figures/hidden-states.png}
%    \caption{\textbf{Performance of the proposed method using HMMs of different order by varying the number of hidden states}.}
%    \label{fig:hidden-states}
%\end{figure}
%
%\begin{table}[!htb]
%\centering
%\caption{PPC codes with the highest probability associated with different states of an HMM trained on successful interviews.}
%\label{tab:emission_matrix_s}
%  \begin{tabular}{|l|l|}
%  \hline
%   \textbf{State} & \textbf{Most likely PPC codes} \\\hline 
%1 & SS, EA, R-CT+, OQ-ECT+, SUM-S, AF, R-CML+ \\\hline           
%2 & LUP+, EA, OQ-ECT+, OQ-ECML+  \\\hline    
%  \end{tabular}
%\end{table} 
%
%\begin{table}[!htb]
%\centering
%\caption{PPC codes with the highest probability associated with different states of an HMM trained on unsuccessful interviews.}
%\label{tab:emission_matrix_u}
%  \begin{tabular}{|l|l|}
%  \hline
%   \textbf{State} & \textbf{Most likely PPC codes} \\\hline     
%1 & OQ-TBN, EA, R-CT+, AR, SUM-S, RBA-S, G-INFO+, R-CML+, OQ-ECT- \\\hline
%2 & LUP+, OQ-ECT+, G-INFO+, CQ-EF, CHT-, HUP-O \\\hline
%  \end{tabular}
%\end{table}
